---
title: "Data visualisation"
author: "Ihsan Khaliq"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data visualisation}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
knitr::opts_chunk$set(progress = TRUE, verbose = TRUE)
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 9,
  fig.height = 9,
  fig.align = "center"
)
```

Load libraries

```{r message=FALSE, echo=TRUE, warning=FALSE}
library("grDevices")
library("tidyverse")
library("viridis")
library("showtext")
library("here")
library("ggpubr")
library("extrafont")
library("kableExtra")
library("tidyr")
library("lubridate")
library("readxl")
library("ggthemes")
library("ggbeeswarm")

theme_set(theme_pubclean(base_family = "Arial", base_size = 14))
```


```{r fonts, include=FALSE, message=FALSE, eval=FALSE, message=FALSE}
# Note that GhostScript needs to be installed at the system level for the PS
# files to be generated. MacOS users can use `brew install ghostscript`
#
# Windows users can follow these directions: 1.	Go to the GhostScript website
# (https://www.ghostscript.com/download/gsdnld.html) 2.	Download the windows
# installer suitable for your machine 3.	Run the installer file which you
# downloaded and follow the prompts 4.	After running the installer click the
# windows "Start" button and type "Edit environment variables for your account"
# and open 5.	In the tab 'Advanced' click the button at the bottom 'Environment
# Variables...' 6.	Under 'System variables' find the variable 'Path', select
# 'Path' and click the 'Edit' button 7. 	Select a new line and copy the
# Ghostscript 'bin' folder location into the field. 7.1	If you installed
# ghostscript to the default folder location; then the folder location will
# likely be "C:\Program Files\gs\gs9.52\bin", the version number (9.52) may
# differ. 8.	Save and exit the environmental variables window This chunk is then
# run only if knitting on new computer that the files have not been generated on
# this is necessary to embed fonts in .eps files
library("extrafont")
if (.Platform$OS.type == "windows") {
   font_import(pattern = "arial", prompt = FALSE)
   loadfonts(device = "postscript", quiet = TRUE)
} else {
   font_import(pattern = "Arial", prompt = FALSE)
   loadfonts(device = "postscript", quiet = TRUE)
}
```



### Import disease data 

```{r import-seasonal-assessments-data}
dt_seasonal <-
  read_excel("data/fungicides_data.xlsx", sheet = 2) %>%
  mutate(across(where(is.character), as.factor)) %>%
  mutate(trial_duration = as.integer(trial_duration)) %>%
  select(
    assessment_date,
    assessment_num,
    year,
    avg_severity_plot,
    avg_severity_treatment,
    diseased_plants_plot,
    perc_incidence_plot,
    perc_incidence_treatment,
    trial_duration
  ) %>%
  dplyr::mutate(severity = avg_severity_treatment / 100) %>%
  dplyr::mutate(disease_severity = avg_severity_plot / 100)
```



### Import weather data

Import weather data and average weather variables by growing season.

```{r import-seasonal-weather-data}
# Calculations over assessment periods
weather_dt_seasonal <-
  read_excel("data/fungicides_data_dailyweather_suffruticosaonly.xlsx",
             sheet = 1) %>%
  dplyr::mutate(assessment_num = as.factor(assessment_num)) %>%
  mutate_if(is.character, as.factor) %>%
  group_by(year) %>%
  summarise(
    min_rh = round(mean(min_rh), 1),
    max_rh = round(mean(max_rh), 1),
    mean_rh = round(mean(mean_rh), 1),
    min_temp = round(mean(min_temp), 1),
    max_temp = round(mean(max_temp), 1),
    mean_temp = round(mean(mean_temp), 1),
    diurnal_temp = round(max_temp - min_temp, 1),
    min_ws = round(mean(min_ws), 1),
    max_ws = round(mean(max_ws), 1),
    mean_ws = round(mean(mean_ws), 1),
    mean_rain = round(mean(precipitation), 1),
    total_rain = round(sum(precipitation), 1),
    total_days = length(unique(date)),
    precip_days = sum(precipitation == "0"),
    # rainless days
    rainy_days = total_days - precip_days,
    perc_rainy_days = rainy_days / total_days * 100,
    avg_daily_rain = round(total_rain / rainy_days, 1),
    rain_everyday = total_rain / total_days
  )

# Filter rainy days to calculate wind speed and temp during rainy days.

weather_dat_rain <-
  read_excel("data/fungicides_data_dailyweather_suffruticosaonly.xlsx",
             sheet = 1) %>%
  select(
    !c(
      assessment_num_orig,
      assessment_num,
      cultivar,
      mean_rh,
      min_rh,
      max_rh,
      mean,
      min,
      max
    )
  ) %>%
  mutate_if(is.character, as.factor) %>%
  filter(precipitation > 0) %>%
  group_by(year) %>%
  summarise(min_temp_rain = round(mean(min_temp), 1),
            max_ws_rain = round(mean(max_ws), 1))

# Filter rainless days to calculate RH during rainless days.

weather_dat_no_rain <-
  read_excel("data/fungicides_data_dailyweather_suffruticosaonly.xlsx",
             sheet = 1) %>%
  select(mean_rh, min_rh, max_rh, precipitation, year) %>%
  mutate_if(is.character, as.factor) %>%
  filter(precipitation == 0) %>%
  group_by(year) %>%
  summarise(max_rh_dry = round(mean(max_rh), 1),
            mean_rh_dry = round(mean(mean_rh), 1))

# Combine data weather data during rainy and rain less periods

combined <-
  left_join(weather_dat_rain, weather_dat_no_rain, by = "year")


# Sort weather data for RH > 65 & below 65%
weather_dat_exposure <-
  read_excel("data/fungicides_data_dailyweather_suffruticosaonly.xlsx",
             sheet = 1) %>%
  select(min_rh, year) %>%
  mutate_if(is.character, as.factor) %>%
  dplyr::group_by(year) %>%
  dplyr::mutate(humidity_catog = as.factor(case_when(min_rh < 65 ~ "RH1",
                                                     min_rh >= 65 ~ "RH2")))

# Count each RH category
weather_dat_exposure_rh <- weather_dat_exposure %>%
  group_by(year, humidity_catog) %>%
  summarize(humidity_count = n()) %>%
  pivot_wider(names_from = humidity_catog, values_from = humidity_count)

# Combine with weather data
combined_rh <-
  left_join(combined,  weather_dat_exposure_rh, by = "year")

# Combine data weather data during rainy and rain less periods with weather recorded over entire duration
weather_dat_seasonal <-
  left_join(weather_dt_seasonal, combined_rh, by = "year")
```


### Cobmine data

Combine seasonal disease assessment data with the corresponding weather data averaged by growing season.

```{r combine-seasonal-data}
dat_seasonal <-
  left_join(dt_seasonal, weather_dat_seasonal, by = "year") %>%
  select(!c(assessment_date,  assessment_num, trial_duration)) %>%
  select(year, everything())
```


Check the data and export the table to excel

```{r kable-data-by-season}
kable(dat_seasonal,
      format = "html",
      table.attr = "class='table table-hover'")
```



### Kernel density plot for disease severity

Kernel density plot to visualize the distribution of disease severity. 

```{r data-distribution-severity, fig.cap="Density plot showing average severity over assessment periods.", message=FALSE, warning=FALSE}
ggplot(dat_seasonal, aes(x = disease_severity)) +
  geom_density(fill = "steelblue", alpha = 0.7) +
  geom_rug(alpha = 0.4) +
  xlab("Disease severity")
```



### Box plot (Mean rain per rainy day - by growing season)

Box plot showing the interaction between average severity, total rainfall and the growing season of the year

```{r boxplot-plot, fig.cap="Box plot showing the relationship between growing season of the year, mean rainfall per rainy day and disease severity.", message=FALSE, warning=FALSE}
fig_1 <- ggplot(
  dat_seasonal,
  aes(
    x = as.factor(year),
    y = disease_severity,
    group = as.factor(year),
    colour = avg_daily_rain,
    fill = avg_daily_rain
  )
) +
  geom_boxplot(alpha = 0.4,
               outlier.size = 0) +
  geom_point(size = 5,
             position = position_jitterdodge()) +
  scale_colour_viridis_c(direction = -1,
                         name = "Mean rain per rainy day (mm)") +
  scale_fill_viridis_c(direction = -1,
                       name = "Mean rain per rainy day (mm)") +
  scale_y_continuous(breaks = seq(from = 0, to = 1, by = 0.25),
                     limits = c(0, 1)) +
  labs(x = "Year",
       y = "Disease severity") +
  guides(color = guide_colorbar(
    title.position = "top",
    title.hjust = 0.5,
    barwidth = unit(20, "lines"),
    barheight = unit(0.5, "lines")
  )) +
  theme(legend.key = element_blank(),
        plot.margin = margin(25, 25, 10, 25)) +
  coord_cartesian(clip = "off",
                  expand = FALSE)
fig_1
```



```{r save-fig-1, include=FALSE, eval=FALSE}
ggsave(
  here("man", "figures/fig_1.png"),
  plot = fig_1,
  width = 9,
  height = 9,
  units = "in",
  dpi = 600
)
ggsave(
  here("man", "figures/fig_1.eps"),
  plot = fig_1,
  width = 9,
  height = 9,
  units = "in",
  dpi = 600,
  device =  cairo_ps
)
```



### Scatter plot (Mean rain per rainy day and wind speed interactions over growing seasons)

```{r plot-severity-rain-wind-speed-season, fig.cap="Scatter plot showing interactions between mean rain per rainy day and daily mean maximum wind speed over growing seasons.", message=FALSE, warning=FALSE}
fig_S2 <- ggplot(dat_seasonal,
                 aes(x = avg_daily_rain, y = disease_severity, color = max_ws)) +
  geom_point(size = 5,
             position = position_jitterdodge()) +
  scale_colour_viridis_c(direction = -1,
                         name = "Daily mean maximum wind speed (m/s)") +
  scale_fill_viridis_c(direction = -1,
                       name = "Daily mean maximum wind speed (m/s)") +
  scale_x_continuous(breaks = seq(2.8, 6.5, length.out = 6),
                     labels = sprintf("%0.1f", seq(2.8, 6.5, length.out = 6))) +
  scale_y_continuous(breaks = seq(from = 0, to = 1, by = 0.25),
                     limits = c(0, 1)) +
  labs(x = "Mean rain per rainy day (mm)", y = "Disease severity") +
  guides(color = guide_colorbar(
    title.position = "top",
    title.hjust = 0.5,
    barwidth = unit(20, "lines"),
    barheight = unit(0.5, "lines")
  )) +
  theme(legend.key = element_blank(),
        plot.margin = margin(25, 25, 10, 25)) +
  coord_cartesian(clip = "off",
                  expand = FALSE)
fig_S2 
```



```{r save-fig-S2, include=FALSE, eval=FALSE}
ggsave(
  here("man", "figures/fig_S2.png"),
  plot = fig_S2,
  width = 9,
  height = 9,
  units = "in",
  dpi = 600
)
ggsave(
  here("man", "figures/fig_S2.eps"),
  plot = fig_S2,
  width = 9,
  height = 9,
  units = "in",
  dpi = 600,
  device =  cairo_ps
)
```



### Scatter plot (average relative humidity and temperature interactions by growing seasons)

```{r scatter-plot-temp-humid-temp-season, fig.cap=" scatter plot showing interactions between daily mean minimum relative humidity and daily mean minimum temperature over growing seasons.", message=FALSE, warning=FALSE}
fig_S4 <- ggplot(dat_seasonal,
                 aes(x = min_temp, y = disease_severity, color = min_rh)) +
  geom_point(size = 5,
             position = position_jitterdodge()) +
  scale_colour_viridis_c(direction = -1,
                         name = "Daily mean minimum relative humidity (%)") +
  scale_fill_viridis_c(direction = -1,
                       name = "Daily mean minimum relative humidity (%)") +
  scale_x_continuous(breaks = seq(7.8, 12.5, length.out = 6),
                     labels = sprintf("%0.1f", seq(7.8, 12.5, length.out = 6))) +
  scale_y_continuous(breaks = seq(from = 0, to = 1, by = 0.25),
                     limits = c(0, 1)) +
  labs(x = "Daily mean minimum temperature (°C)", y = "Disease severity") +
  guides(color = guide_colorbar(
    title.position = "top",
    title.hjust = 0.5,
    barwidth = unit(20, "lines"),
    barheight = unit(0.5, "lines")
  )) +
  theme(legend.key = element_blank(),
        plot.margin = margin(25, 25, 10, 25)) +
  coord_cartesian(clip = "off",
                  expand = FALSE)
fig_S4
```


```{r save-fig-S4, include=FALSE, eval=FALSE}
ggsave(
  here("man", "figures/fig_S4.png"),
  plot = fig_S4,
  width = 9,
  height = 9,
  units = "in",
  dpi = 600
)
ggsave(
  here("man", "figures/fig_S4.eps"),
  plot = fig_S4,
  width = 9,
  height = 9,
  units = "in",
  dpi = 600,
  device =  cairo_ps
)
```


## Plot weather data for the entire 9-year duration 

### Import data

```{r}
daily_df <-
  read_excel("data/dat_nine_years.xlsx",
             sheet = 3) %>% 
  mutate(julian_day = yday(date)) %>%
  arrange(month(date), day(date), year, julian_day, date) %>%
  relocate(julian_day, .after = date) %>%
  group_by(julian_day) %>%
  summarise(
    min_rh = round(mean(min_rh), 1),
    max_rh = round(mean(max_rh), 1),
    mean_rh = round(mean(mean_rh), 1),
    min_temp = round(mean(min_temp), 1),
    max_temp = round(mean(max_temp), 1),
    mean_temp = round(mean(mean_temp), 1),
    max_ws = round(mean(max_ws), 1),
    mean_rain = round(mean(precipitation), 1),
    total_rain = round(sum(precipitation), 1)
  )
```


### Plot temperature data for the entire 9-year duration 

```{r}
temp_df <- daily_df %>%
  pivot_longer(
    cols = c("min_temp", "max_temp", "mean_temp"),
    names_to = "temp_type",
    values_to = "value"
  )

p1 <- ggplot(temp_df, aes(julian_day, value, color = temp_type)) +
   geom_point(position = position_jitterdodge(jitter.width = 0.01), size = 2) +
  scale_y_continuous(breaks = seq(from = -3.2, to = 28, by = 6), limits = c(-3.2, 28)) +
    theme_few() + 
  labs(x = "Julian day", y = "Temperature (°C)", color = "Temperature type") 

p1
```


### Plot RH data for the entire 9-year duration 


```{r}
rh_df <- daily_df %>%
  pivot_longer(
    cols = c("min_rh", "max_rh", "mean_rh"),
    names_to = "rh_type",
    values_to = "value"
  )

p2 <- ggplot(rh_df, aes(julian_day, value, color = rh_type)) +
   geom_point(position = position_jitterdodge(jitter.width = 0.01), size = 2) +
  scale_y_continuous(breaks = seq(from = 0, to = 100, by = 25), limits = c(0, 100)) +
    theme_few() + 
  labs(x = "Julian day", y = "Relative humidity (%)", color = "RH type") 

p2
```






