---
title: "GAMs"
author: "Ihsan Khaliq"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GAMs}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
knitr::opts_chunk$set(progress = TRUE, verbose = TRUE)
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 9,
  fig.height = 9,
  fig.align = "center"
)
```


Load libraries

```{r message=FALSE, echo=TRUE, warning=FALSE}
library("readxl")
library("tidyverse")
library("ggeffects")
library("here")
library("lubridate")
library("ggpubr")
library("mgcv")
library("kableExtra")
library("DHARMa")
library("gratia")
library("mgcViz")
library("patchwork")
library("broom")

# theme_set(theme_pubr(base_family = "Arial", base_size = 15))
```



### Import disease data

```{r import-seasonal-assessments-data}
dt_seasonal <-
  read_excel("data/fungicides_data.xlsx", sheet = 2) %>%
  mutate(across(where(is.character), as.factor)) %>%
  mutate(trial_duration = as.integer(trial_duration)) %>%
  select(
    assessment_date,
    assessment_num,
    year,
    avg_severity_plot,
    avg_severity_treatment,
    diseased_plants_plot,
    perc_incidence_plot,
    perc_incidence_treatment,
    trial_duration
  ) %>%
  dplyr::mutate(severity = avg_severity_treatment / 100) %>%
  dplyr::mutate(disease_severity = avg_severity_plot / 100)
```



### Import weather data

```{r import-seasonal-weather-data}
# Calculations over assessment periods
weather_dt_seasonal <-
  read_excel("data/fungicides_data_dailyweather_suffruticosaonly.xlsx",
             sheet = 1) %>%
  dplyr::mutate(assessment_num = as.factor(assessment_num)) %>%
  mutate_if(is.character, as.factor) %>%
  group_by(year) %>%
  summarise(
    min_rh = round(mean(min_rh), 1),
    max_rh = round(mean(max_rh), 1),
    mean_rh = round(mean(mean_rh), 1),
    min_temp = round(mean(min_temp), 1),
    max_temp = round(mean(max_temp), 1),
    mean_temp = round(mean(mean_temp), 1),
    diurnal_temp = round(max_temp - min_temp, 1),
    min_ws = round(mean(min_ws), 1),
    max_ws = round(mean(max_ws), 1),
    mean_ws = round(mean(mean_ws), 1),
    mean_rain = round(mean(precipitation), 1),
    total_rain = round(sum(precipitation), 1),
    total_days = length(unique(date)),
    precip_days = sum(precipitation == "0"),
    # rainless days
    rainy_days = total_days - precip_days,
    perc_rainy_days = rainy_days / total_days * 100,
    avg_daily_rain = round(total_rain / rainy_days, 1),
    rain_everyday = total_rain / total_days
  )

# Filter rainy days to calculate wind speed and temp during rainy days.

weather_dat_rain <-
  read_excel("data/fungicides_data_dailyweather_suffruticosaonly.xlsx",
             sheet = 1) %>%
  select(
    !c(
      assessment_num_orig,
      assessment_num,
      cultivar,
      mean_rh,
      min_rh,
      max_rh,
      mean,
      min,
      max
    )
  ) %>%
  mutate_if(is.character, as.factor) %>%
  filter(precipitation > 0) %>%
  group_by(year) %>%
  summarise(min_temp_rain = round(mean(min_temp), 1),
            max_ws_rain = round(mean(max_ws), 1))

# Filter rainless days to calculate RH during rainless days.

weather_dat_no_rain <-
  read_excel("data/fungicides_data_dailyweather_suffruticosaonly.xlsx",
             sheet = 1) %>%
  select(mean_rh, min_rh, max_rh, precipitation, year) %>%
  mutate_if(is.character, as.factor) %>%
  filter(precipitation == 0) %>%
  group_by(year) %>%
  summarise(max_rh_dry = round(mean(max_rh), 1),
            mean_rh_dry = round(mean(mean_rh), 1))

# Combine data weather data during rainy and rain less periods

combined <-
  left_join(weather_dat_rain, weather_dat_no_rain, by = "year")


# Sort weather data for RH > 65 & below 65%
weather_dat_exposure <-
  read_excel("data/fungicides_data_dailyweather_suffruticosaonly.xlsx",
             sheet = 1) %>%
  select(min_rh, year) %>%
  mutate_if(is.character, as.factor) %>%
  dplyr::group_by(year) %>%
  dplyr::mutate(humidity_catog = as.factor(case_when(min_rh < 65 ~ "RH1",
                                                     min_rh >= 65 ~ "RH2")))

# Count each RH category
weather_dat_exposure_rh <- weather_dat_exposure %>%
  group_by(year, humidity_catog) %>%
  summarize(humidity_count = n()) %>%
  pivot_wider(names_from = humidity_catog, values_from = humidity_count)

# Combine with weather data
combined_rh <-
  left_join(combined,  weather_dat_exposure_rh, by = "year")

# Combine data weather data during rainy and rain less periods with weather recorded over entire duration
weather_dat_seasonal <-
  left_join(weather_dt_seasonal, combined_rh, by = "year")
```



### Combine data

```{r combine-seasonal-data}
dat_seasonal <-
  left_join(dt_seasonal, weather_dat_seasonal, by = "year") %>%
  select(!c(assessment_date,  assessment_num, trial_duration)) %>%
  select(year, everything())
```



### Set seed for reproducibility

```{r}
set.seed(42)
```



# Main effect of weather variables


## M1 (min_rh, min_temp, max_ws, mean_rain_per_rainy_day)

Using mean rain per rainy day, and mean values for daily minimum relative
humidity, daily minimum temperatures, and daily maximum wind speed  over the trial duration.

```{r m1}
m1 <- gam(disease_severity ~  s(min_rh, k = 10) + s(min_temp, k = 10) + s(max_ws, k =7) + s(avg_daily_rain, k = 9) + s(total_days, k=9), family = betar(),  method = "REML", data = dat_seasonal)

summary(m1)

draw(m1, resid_col = TRUE, rug = FALSE, caption = NULL)
```



```{r, include=FALSE, eval=FALSE}
# Customise plot using gratia package
p1 <- draw(m1, select = "s(min_rh)", rug = FALSE, ) +
  labs(x = "Mean minimum RH (%)", title = "", caption = NULL) +
  scale_x_continuous(breaks = seq(29.7, 67.9, length.out = 6),
                     labels = sprintf("%0.1f", seq(29.7, 67.9, length.out = 6))) + theme_bw()

p2 <- draw(m1, select = "s(min_temp)", rug = FALSE) +
  labs(
    x = "Mean minimum temperature (°C)",
    y = "",
    title = "",
    caption = NULL
  ) +
  scale_x_continuous(breaks = seq(7.8, 12.5, length.out = 6),
                     labels = sprintf("%0.1f", seq(7.8, 12.5, length.out = 6))) + theme_bw()


p3 <- draw(m1, select = "s(max_ws)", rug = FALSE) +
  labs(
    x = "Mean minimum wind speed (m/s)",
    y = "",
    title = "",
    caption = NULL
  ) +
  scale_x_continuous(breaks = seq(2.8, 3.7, length.out = 5),
                     labels = sprintf("%0.1f", seq(2.8, 3.7, length.out = 5))) + theme_bw()


p4 <- draw(m1, select = "s(avg_daily_rain)", rug = FALSE) +
  labs(x = "Mean rain per rainy day (mm)", title = "", caption = NULL) +
  scale_x_continuous(breaks = seq(2.8, 6.5, length.out = 6),
                     labels = sprintf("%0.1f", seq(2.8, 6.5, length.out = 6))) + theme_bw()
p5 <- draw(m1, select = "s(total_days)", rug = FALSE) +
  labs(
    x = "Trial duration (days)",
    y = "",
    title = "",
    caption = NULL
  ) +
  scale_x_continuous(breaks = seq(18, 200, length.out = 6),
                     labels = as.character(as.integer(seq(18, 200, length.out = 6)))) + theme_bw()
fig_2 <- p1 + p2 + p3 + p4 + p5 + plot_layout(ncol = 3, nrow = 2)
```



```{r save-fig2, include=FALSE, eval=FALSE}
ggsave(
  here("man", "figures/fig_2.png"),
  plot = fig_2,
  width = 9,
  height = 9,
  units = "in",
  dpi = 600
)
ggsave(
  here("man", "figures/fig_2.eps"),
  plot = fig_2,
  width = 9,
  height = 9,
  units = "in",
  dpi = 600,
  device =  cairo_ps
)
```


## M2 (min_rh, min_temp_rain, max_ws_rain, mean_rain_per_rainy_day)

Based on correlation analysis, daily minimum temperature and daily maximum wind speed during rainy days has a higher correlation with disease, so use these variables instead of mean values for daily minimum temperatures and daily maximum wind speed over the trial duration.

```{r m2}
m2 <- gam(disease_severity ~  s(min_rh, k = 9) + s(min_temp_rain, k = 9) + s(max_ws_rain, k =7) + s(avg_daily_rain, k = 9) + s(total_days, k=9), family = betar(),  method = "REML", data = dat_seasonal)

summary(m2)

draw(m2, resid_col = TRUE, rug = FALSE, caption = NULL)
```


Same output as `m1`.



## M3 (max_rh_dry, min_temp_rain, max_ws_rain, mean_rain_per_rainy_day)

Add daily maximum relative humidity during dry days, which has the highest
correlation with the disease.

```{r m3}
m3 <- gam(disease_severity ~  s(max_rh_dry, k = 10) + s(min_temp_rain, k = 9) + s(max_ws_rain, k =7) + s(avg_daily_rain, k = 9) + s(total_days, k=10), family = betar(),  method = "REML", data = dat_seasonal)

summary(m3)

draw(m3, resid_col = TRUE, rug = FALSE, caption = NULL)
```


Only rainfall has a significant on disease severity. 



## M4 (RH2, min_temp_rain, max_ws_rain, mean_rain_per_rainy_day)

Use the number of days with relative humidity over 65% as a predictor, which has the second highest correlation with the disease after after maximum relative humidity during dry conditions.

```{r m4}
m4 <- gam(disease_severity ~  s(RH2, k = 9) + s(min_temp_rain, k = 9) + s(max_ws_rain, k =7) + s(avg_daily_rain, k = 9) + s(total_days, k=9), family = betar(),  method = "REML", data = dat_seasonal)

summary(m4)

draw(m4, resid_col = TRUE, rug = FALSE, caption = NULL)
```


Same output as `m1` and `m2`, but trial duration has a significant effect. This means trial duration is important when number of days with relative humidity over 65% is used as a predictor, so this predictor is not normalising the data for varying trial period.



## M5 (Mean RH, min_temp_rain, max_ws_rain, mean_rain_per_rainy_day)

Use mean RH as a predictor, which has the third highest correlation with the disease after after maximum relative humidity during dry conditions. 

```{r m5}
m5 <- gam(disease_severity ~  s(mean_rh, k = 9) + s(min_temp, k = 9) + s(max_ws, k =7) + s(avg_daily_rain, k = 10) + s(total_days, k=9), family = betar(),  method = "REML", data = dat_seasonal)

summary(m5)

draw(m5, resid_col = TRUE, rug = FALSE, caption = NULL)
```


Only rain has no significant effect. Trial duration has a significant effect. This means trial duration is important when number of days with relative humidity over 65% is used as a predictor, so this predictor is not normalising the data for varying trial period.



## Compare models and choose the best fit model 

### AIC, BIC 

```{r}
models <- list(
  mod1 = m1,
  mod2 = m2,
  mod3 = m3,
  mod4 = m4,
  mod5 = m5
)
map_df(models, glance, .id = "model") %>%
  arrange(AIC)
```

The `m1` model has lowest AIC and BIC values

### R-squared values

```{r}
enframe(
  c(
    mod1 = summary(m1)$r.sq,
    mod2 = summary(m2)$r.sq,
    mod3 = summary(m3)$r.sq,
    mod4 = summary(m4)$r.sq,
    mod5 = summary(m5)$r.sq
  )
) %>%
  arrange(desc(value))
```

The `m1` model has the highest R-squared values. 

## Best fit model `m` diagnostics

### Check if the best fit model `m1` predictions are reliable 

```{r check-m1}
simulateResiduals(m1, plot = T, quantreg = T)
```

Yes, predictions are reliable and truly represent the data.



### Diagnostics plots

```{r diagnostics-plot-m1}
m1_vis <- getViz(m1)
check(
  m1_vis,
  a.qq = list(method = "tnorm",
              a.cipoly = list(fill = "light blue")),
  a.respoi = list(size = 0.5),
  a.hist = list(bins = 10)
)
```


The `k` values are greater than `edf` and p-values are not significant. More data will further improve the fit.



### Check autocorrelation

```{r check-autocorrelation}
performance::check_autocorrelation(m1)
```



# Interaction effect of weather variables

```{r mod1}
dat_season <- dat_seasonal %>%
  rename(rain_per_rainy_day = avg_daily_rain,
         daily_minimum_temperature = min_temp)

mod1 <-
  gam(
    disease_severity ~  te(min_rh,  daily_minimum_temperature, k = 4) +  te(max_ws, rain_per_rainy_day, k = 5),
    family = betar(),
    method = "REML",
    data = dat_season
  )

summary(mod1)

p1 <- draw(mod1,  rug = FALSE,  select = smooths(mod1)[1]) +
  labs(
    x = "Daily mean minimum RH (%)",
    y = "Daily mean minimum temperature (°C)",
    title = "Temperature and RH interactions",
    caption = NULL
  ) +
  scale_y_continuous(breaks = seq(7.8, 12.5, length.out = 6),
                     labels = sprintf("%0.1f", seq(7.8, 12.5, length.out = 6)))

p2 <- draw(mod1,  rug = FALSE,  select = smooths(mod1)[2]) +
  labs(
    x = "Daily mean maximum wind speed (m/s)",
    y = "Mean rain per rainy day (mm)",
    title = "Wind speed and rain interactions",
    caption = NULL
  ) +
  scale_y_continuous(breaks = seq(2.9, 6.5, length.out = 5),
                     labels = sprintf("%0.1f", seq(2.9, 6.5, length.out = 5)))


fig_3 <- p1 + p2 + plot_layout(ncol = 2)
fig_3 
```



```{r save-fig3, include=FALSE, eval=FALSE}
ggsave(
  here("man", "figures/fig_3.png"),
  plot = fig_3,
  width = 9,
  height = 9,
  units = "in",
  dpi = 600
)
ggsave(
  here("man", "figures/fig_3.eps"),
  plot = fig_3,
  width = 9,
  height = 9,
  units = "in",
  dpi = 600,
  device =  cairo_ps
)
```



## Best fit model `m2` diagnostics

### Check if the model predictions are reliable 

```{r simulate-residuals}
simulateResiduals(mod1, plot = T, quantreg = T)
```

No problems detected. 


### Diagnostics plots

```{r diagnostics-plot-mod1}
mod1_vis <- getViz(m1)
check(
  mod1_vis,
  a.qq = list(method = "tnorm",
              a.cipoly = list(fill = "light blue")),
  a.respoi = list(size = 0.5),
  a.hist = list(bins = 10)
)
```

The `k` values are greater than `edf` and p-values are not significant. 


### Check autocorrelation

```{r mod1-autocorrelation}
performance::check_autocorrelation(mod1)
```



### Plot interaction effect of mean rain per rainy day and daily mean maximum
wind speed using `visreg` package

```{r fig_S2}
fig_S2 <-
  visreg::visreg(
    mod1,
    "max_ws",
    by = "rain_per_rainy_day",
    gg = TRUE,
    rug = FALSE,
    partial = FALSE,
    line = list(col = "black")
  ) +  theme_bw() +
  labs(x = "Daily mean maximum wind speed (m/s)", y = "Partial effect") +
  scale_x_continuous(breaks = seq(2.8, 3.7, length.out = 5),
                     labels = sprintf("%0.1f", seq(2.8, 3.7, length.out = 5)))


fig_S2 
```



```{r save-figS2, include=FALSE, eval=FALSE}
ggsave(
  here("man", "figures/fig_S2.png"),
  plot = fig_S2,
  width = 9,
  height = 9,
  units = "in",
  dpi = 600
)
ggsave(
  here("man", "figures/fig_S2.eps"),
  plot = fig_S2,
  width = 9,
  height = 9,
  units = "in",
  dpi = 600,
  device =  cairo_ps
)
```



### Plot interaction effect of mean mininum RH and mean mininum temp

```{r}
fig_S3 <-
  visreg::visreg(
    mod1,
    "min_rh",
    by = "daily_minimum_temperature",
    gg = TRUE,
    rug = FALSE,
    partial = FALSE,
    line = list(col = "black")
  ) + theme_bw() +
  labs(x = "Daily mean minimum RH (%)",  y = "Partial effect") +
  scale_x_continuous(breaks = seq(29.7, 67.9, length.out = 6),
                     labels = sprintf("%0.1f", seq(29.7, 67.9, length.out = 6)))

fig_S3   
```


The plot for interaction effect between RH and temperature suggests a negative interaction effect. 

```{r save-figS3, include=FALSE, eval=FALSE}
ggsave(
  here("man", "figures/fig_S3.png"),
  plot = fig_S3,
  width = 9,
  height = 9,
  units = "in",
  dpi = 600
)
ggsave(
  here("man", "figures/fig_S3.eps"),
  plot = fig_S3,
  width = 9,
  height = 9,
  units = "in",
  dpi = 600,
  device =  cairo_ps
)
```


The plot for the interaction effect between max_ws and avg_daily_rain suggests a negative interaction effect. 



```{r, include=FALSE}
# Table listing number of days below zero, and minimum & maximum temperatures recorded from planting date to the date to November 1 of the previous year.
dat_weather <-
  read_excel(
    "data/fungicides_data_dailyweather_suffruticosaonly.xlsx",
    sheet = 3,
    na = "NA"
  ) %>%
  select(!c(mean, min, max, number)) %>%
  dplyr::mutate(year_ps = as.factor(year_ps)) %>%
  dplyr::mutate(across(where(is.character), as.numeric)) %>%
  group_by(year_ps) %>%
  summarise(
    total_days = length(unique(date)),
    temp_days_0 = sum(min_temp < 0, na.rm = TRUE),
    perc_temp_days_0 = temp_days_0 / total_days * 100,
    mean_min_temp = mean(min_temp, na.rm = TRUE),
    mean_max_temp = mean(max_temp, na.rm = TRUE),
    min_temp = min(min_temp, na.rm = TRUE),
    max_temp = max(max_temp, na.rm = TRUE),
  )

# openxlsx::write.xlsx(dat_weather , "Threshold table.xlsx", rowNames = FALSE)
```









